---
title: "biwhitening PMF"
author: "DongyueXie"
date: "2022-03-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

One of the feature we'd like to add to Topic model is to induce sparsity on loadings, and factors, for better interpretability.

Directly adding such constraint is non-trivial, and has not been shown to perform well, especialy using empirical Bayes approach.

Here I took an approach of combining [biwhitening](biwhitening.html), and EBNMF(a scaled version, and I wrote the code and is available in `funflash` package). In this analysis, we show how the method performs well on simple examples, and can acutally induce sparsity on L, and F.

The examples are taken from [here](https://stephens999.github.io/misc/nmf_sparse.html)

```{r}
library(funflash)
```

## Simple example, K=3

This is an example where topic model can give sparse estimates, because of sparsity of the data.

The biwhitening + ebnmf method can also recover the true structure. (although it gives slightly more dimensions, the true ones are well recovered, and the redundant ones have very small PVE.) Note also that we input the true K to topic model.

```{r}
set.seed(12345)
n = 99
p = 300
k= 3
L = matrix(0, nrow=n, ncol=k)
F = matrix(0, nrow=p, ncol=k)

L[1:(n/3),1] = 1
L[((n/3)+1):(2*n/3),2] = 1
L[((2*n/3)+1):n,3] = 1

F[1:(p/3),1] = 1+10*runif(p/3)
F[((p/3)+1):(2*p/3),2] = 1+10*runif(p/3)
F[((2*p/3)+1):p,3] = 1+10*runif(p/3)

lambda = L %*% t(F)
X = matrix(rpois(n=length(lambda),lambda),nrow=n)
image(X)

fit_tm = fastTopics::fit_poisson_nmf(X,3)
plot(fit_tm$L[,1],main="Topic model, estimated loadings 1")
plot(fit_tm$L[,2],main="Topic model, estimated loadings 2")
plot(fit_tm$L[,3],main="Topic model, estimated loadings 3")

Y_tilde = biwhitening(X)
fit_sf = scaledflash(Y_tilde$Y,Y_tilde$u,Y_tilde$v,
                     S2 = NULL,
                     var.type = 'by_column',
                     Kmax=10,
                     tol=0.01,
                     maxiter = 1000,
                     ebnm_fn = 'ebnm_pe',
                     init_fn = 'nnmf_r1',
                     ebnm_param=NULL,
                     verbose=TRUE,
                     nullcheck=TRUE,
                     sigma2 = NULL,
                     seed=12345)
plot(fit_sf$ldf$l[,1],main="biwhitening+flash, estimated loadings 1")
plot(fit_sf$ldf$l[,2],main="biwhitening+flash, estimated loadings 1")
plot(fit_sf$ldf$l[,3],main="biwhitening+flash, estimated loadings 1")

fit_sf$pve
```

## Harder example, K=4

This is a harder example because there's a background now, and we see that topic model cannot give sparse estimates.

However, biwhitening + ebnmf gives the sparse recovery, and also estimates the correct K.

```{r}
#################
set.seed(12345)
n = 99
p = 300
k= 4
mfac = 3 # controls PVE of dense factor
L = matrix(0, nrow=n, ncol=k)
F = matrix(0, nrow=p, ncol=k)

L[1:(n/3),1] = 1
L[((n/3)+1):(2*n/3),2] = 1
L[((2*n/3)+1):n,3] = 1
L[,4] = 1+mfac*runif(n)

F[1:(p/3),1] = 1+10*runif(p/3)
F[((p/3)+1):(2*p/3),2] = 1+10*runif(p/3)
F[((2*p/3)+1):p,3] = 1+10*runif(p/3)
F[,4]= 1+mfac*runif(p)

lambda = L %*% t(F)
X = matrix(rpois(n=length(lambda),lambda),nrow=n)
image(X)
```

```{r}
fit_tm = fastTopics::fit_poisson_nmf(X,4)
plot(fit_tm$L[,1],main="Topic model, estimated loadings 1")
plot(fit_tm$L[,2],main="Topic model, estimated loadings 2")
plot(fit_tm$L[,3],main="Topic model, estimated loadings 3")
plot(fit_tm$L[,4],main="Topic model, estimated loadings 4")
```

```{r}
Y_tilde = biwhitening(X)
fit_sf = scaledflash(Y_tilde$Y,Y_tilde$u,Y_tilde$v,
                     S2 = NULL,
                     var.type = 'by_column',
                     Kmax=10,
                     tol=0.01,
                     maxiter = 1000,
                     ebnm_fn = 'ebnm_pe',
                     init_fn = 'nnmf_r1',
                     ebnm_param=NULL,
                     verbose=TRUE,
                     nullcheck=TRUE,
                     sigma2 = NULL,
                     seed=12345)
plot(fit_sf$ldf$l[,1],main="biwhitening+flash, estimated loadings 1")
plot(fit_sf$ldf$l[,2],main="biwhitening+flash, estimated loadings 2")
plot(fit_sf$ldf$l[,3],main="biwhitening+flash, estimated loadings 3")
plot(fit_sf$ldf$l[,4],main="biwhitening+flash, estimated loadings 4")
```

```{r}
library(RcppML)
library(NMF)
```

```{r}
fitted_rcppml_nmf = function(fit){
  fit$w%*%diag(fit$d)%*%fit$h
}
n_try = 10
fit_nmf1_best = RcppML::nmf(X,4,L1=c(0.99,0.99),seed=0)
error_best = sum((X-fitted_rcppml_nmf(fit_nmf1_best))^2)
for(i in 1:n_try){
  fit_nmf1 = RcppML::nmf(X,4,L1=c(0.99,0.99),seed=i)
  error = sum((X-fitted_rcppml_nmf(fit_nmf1))^2)
  if(error<error_best){
    fit_nmf1_best = fit_nmf1
    error_best = error
  }
}

```

```{r}
fitted_nmf2 = function(fit){
  fit@fit@W%*%fit@fit@H
}
beta = 0.5
fit_nmf2_best = NMF::nmf(X,4,'snmf/l',beta = beta,seed=0)
error_best = sum((X-fitted_nmf2(fit_nmf2_best))^2)
for(i in 1:n_try){
  fit_nmf2 = NMF::nmf(X,4,'snmf/l',beta = beta,seed=i)
  error = sum((X-fitted_nmf2(fit_nmf2))^2)
  if(error<error_best){
    fit_nmf2_best = fit_nmf2
    error_best = error
  }
}
```

```{r}
my_ldf = function(Lhat,Fhat){
  dl = apply(Lhat,2,norm,type='2')
  df = apply(Fhat,2,norm,type='2')
  return(list(l = apply(Lhat,2,function(z){z/norm(z,'2')}),
              f = apply(Fhat,2,function(z){z/norm(z,'2')}),
              d = dl*df))
}
library(ggplot2)
library(gridExtra)
library(grid)
plots = list()
for(i in 1:4){
  temp = my_ldf(L,F)
  data <- data.frame(x=1:99,y =temp$l[,i])
  plots[[i]] <- ggplot(data, aes(x = x,y=y)) + geom_point() + theme(axis.title.x = element_blank(),axis.title.y = element_blank())
}
do.call(grid.arrange,c(plots,list(nrow=2,bottom=textGrob("True L"))))

for(i in 1:4){
  temp = my_ldf(fit_tm$L,fit_tm$F)
  data <- data.frame(x=1:99,y =temp$l[,i])
  plots[[i]] <- ggplot(data, aes(x = x,y=y)) + geom_point() + theme(axis.title.x = element_blank(),axis.title.y = element_blank())
}
do.call(grid.arrange,c(plots,list(nrow=2,bottom=textGrob("fastTopics::fit_poisson_nmf"))))

for(i in 1:4){
  temp = my_ldf(fit_sf$ldf$l,fit_sf$ldf$f)
  data <- data.frame(x=1:99,y =temp$l[,i])
  plots[[i]] <- ggplot(data, aes(x = x,y=y)) + geom_point() + theme(axis.title.x = element_blank(),axis.title.y = element_blank())
}
do.call(grid.arrange,c(plots,list(nrow=2,bottom=textGrob("Biwhitening EBNMF"))))

for(i in 1:4){
  temp = my_ldf(fit_nmf1_best$w,t(fit_nmf1_best$h))
  data <- data.frame(x=1:99,y =temp$l[,i])
  plots[[i]] <- ggplot(data, aes(x = x,y=y)) + geom_point() + theme(axis.title.x = element_blank(),axis.title.y = element_blank())
}
do.call(grid.arrange,c(plots,list(nrow=2,bottom=textGrob("RcppML::nmf"))))

for(i in 1:4){
  temp = my_ldf(fit_nmf2_best@fit@W,t(fit_nmf2_best@fit@H))
  data <- data.frame(x=1:99,y =temp$l[,i])
  plots[[i]] <- ggplot(data, aes(x = x,y=y)) + geom_point() + theme(axis.title.x = element_blank(),axis.title.y = element_blank())
}
do.call(grid.arrange,c(plots,list(nrow=2,bottom=textGrob("NMF::nmf"))))
```


plot F

```{r}
plots = list()
for(i in 1:4){
  temp = my_ldf(L,F)
  data <- data.frame(x=1:300,y =temp$f[,i])
  plots[[i]] <- ggplot(data, aes(x = x,y=y)) + geom_point() + theme(axis.title.x = element_blank(),axis.title.y = element_blank())
}
do.call(grid.arrange,c(plots,list(nrow=2,bottom=textGrob("True F"))))

for(i in 1:4){
  temp = my_ldf(fit_tm$L,fit_tm$F)
  data <- data.frame(x=1:300,y =temp$f[,i])
  plots[[i]] <- ggplot(data, aes(x = x,y=y)) + geom_point() + theme(axis.title.x = element_blank(),axis.title.y = element_blank())
}
do.call(grid.arrange,c(plots,list(nrow=2,bottom=textGrob("fastTopics::fit_poisson_nmf"))))

for(i in 1:4){
  temp = my_ldf(fit_sf$ldf$l,fit_sf$ldf$f)
  data <- data.frame(x=1:300,y =temp$f[,i])
  plots[[i]] <- ggplot(data, aes(x = x,y=y)) + geom_point() + theme(axis.title.x = element_blank(),axis.title.y = element_blank())
}
do.call(grid.arrange,c(plots,list(nrow=2,bottom=textGrob("Biwhitening EBNMF"))))

for(i in 1:4){
  temp = my_ldf(fit_nmf1_best$w,t(fit_nmf1_best$h))
  data <- data.frame(x=1:300,y =temp$f[,i])
  plots[[i]] <- ggplot(data, aes(x = x,y=y)) + geom_point() + theme(axis.title.x = element_blank(),axis.title.y = element_blank())
}
do.call(grid.arrange,c(plots,list(nrow=2,bottom=textGrob("RcppML::nmf"))))

for(i in 1:4){
  temp = my_ldf(fit_nmf2_best@fit@W,t(fit_nmf2_best@fit@H))
  data <- data.frame(x=1:300,y =temp$f[,i])
  plots[[i]] <- ggplot(data, aes(x = x,y=y)) + geom_point() + theme(axis.title.x = element_blank(),axis.title.y = element_blank())
}
do.call(grid.arrange,c(plots,list(nrow=2,bottom=textGrob("NMF::nmf"))))
```

