---
title: "EBPM Scale issue"
author: "Dongyue Xie"
date: "2019-11-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In EBPM, the posterior mean is 
\begin{equation}
\begin{split}
    E(\lambda|x,s) &= \tilde{\pi}_0\frac{x+a}{b+\delta+s}+(1-\tilde{\pi}_0)\frac{x+a}{b+s}
    \\ & = [\tilde{\pi}_0\frac{s}{b+\delta+s}+(1-\tilde{\pi}_0)\frac{s}{b+s}]\frac{x+a}{s},
\end{split}
\end{equation}
where 
\begin{equation}
    \tilde{\pi}_0 = \frac{\pi_0 NB(x;a,\frac{b+\delta}{b+\delta+s})}{\pi_0 NB(x;a,\frac{b+\delta}{b+\delta+s}) + (1-\pi_0) NB(x;a,\frac{b}{b+s})}.
\end{equation}

Plugging in the pmf of Negative Binomial distribution, we have 
\begin{equation}
\tilde{\pi}_0 = \frac{1}{1+\frac{1-\pi_0}{\pi_0}(1+\frac{\delta s}{b(b+\delta+s)})^{-a}(1+\frac{\delta}{b+s})^{x}}.
\end{equation}

```{r}
ebpm_pm = function(a,b,delta,pi0,s,x){
  pi0_pm = pi0*dnbinom(x,a,(b+delta)/(b+delta+s))
  pi1_pm = (1-pi0)*dnbinom(x,a,(b)/(b+s))
  pi0_pm = pi0_pm/(pi0_pm+pi1_pm)
  shrink_factor = pi0_pm*(s/(b+delta+s))+(1-pi0_pm)*(s/(b+s))
  lambda_pm = shrink_factor*((x+a)/s)
  return(list(pi0_pm=pi0_pm,shrink_factor=shrink_factor,lambda_pm=lambda_pm))
}

```

Now assume $b=0.01$, $\delta=100$, $a=1$, $\pi_0=0.8$,

Let $s=1$ and $x$ range from 0 to 20, compare $\tilde{\pi}_0$, and $E(\lambda|y,s)$.

```{r}
a=1
b=0.01
delta=100
pi0=0.8
s=1
x=seq(0,50)
sf=c()
lp=c()
pp=c()
for(i in 1:length(x)){
  fit = ebpm_pm(a,b,delta,pi0,s,x[i])
  sf=c(sf,fit$shrink_factor)
  lp = c(lp,fit$lambda_pm)
  pp = c(pp,fit$pi0_pm)
}
par(mfrow=c(1,2))
#plot(x,sf,main='shrinkage factor',xlab='x',ylab='shrinkage factor',ylim=c(0,1),pch=16)
plot(x,pp,main='Pi0 tilde',xlab='x',ylab='Pi0 tilde',ylim=c(0,1),pch=16)
plot(x,lp,main='Posterior Mean',xlab='x',ylab='Posterior mean of lambda',pch=16)
abline(0,1)
```


Let $s=100$.

```{r}
a=1
b=0.01
delta=100
pi0=0.8
s=100
x=seq(0,100)
sf=c()
lp=c()
pp=c()
for(i in 1:length(x)){
  fit = ebpm_pm(a,b,delta,pi0,s,x[i])
  sf=c(sf,fit$shrink_factor)
  lp = c(lp,fit$lambda_pm)
  pp = c(pp,fit$pi0_pm)
}
par(mfrow=c(1,2))
#plot(x,sf,main='shrinkage factor',xlab='x',ylab='shrinkage factor',ylim=c(0,1),pch=16)
plot(x,pp,main='Pi0 tilde',xlab='x',ylab='Pi0 tilde',ylim=c(0,1),pch=16)
plot(x/s,lp,main='Posterior Mean',xlab='x',ylab='Posterior mean of lambda',pch=16)
abline(0,1)
```

Let $s=300$.

```{r}
a=1
b=0.01
delta=100
pi0=0.8
s=300
x=seq(0,100)
sf=c()
lp=c()
pp=c()
for(i in 1:length(x)){
  fit = ebpm_pm(a,b,delta,pi0,s,x[i])
  sf=c(sf,fit$shrink_factor)
  lp = c(lp,fit$lambda_pm)
  pp = c(pp,fit$pi0_pm)
}
par(mfrow=c(1,2))
#plot(x,sf,main='shrinkage factor',xlab='x',ylab='shrinkage factor',ylim=c(0,1),pch=16)
plot(x,pp,main='Pi0 tilde',xlab='x',ylab='Pi0 tilde',ylim=c(0,1),pch=16)
plot(x/s,lp,main='Posterior Mean',xlab='x',ylab='Posterior mean of lambda',pch=16)
abline(0,1)
```
