---
title: "fixFupdateL in fastTopics"
author: "DongyueXie"
date: "2022-03-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In some cases we would like to fix facrors $F$ and only get $L$ estimated in the following model, $Y\sim Poisson(LF^T)$, where $L\in R_+^{n\times K}$ and $F\in R_{+}^{m\times K}$.

I found that `fastTopics` has the option to fix factors and now let's try this feature.


```{r}
library(fastTopics)
K = 3
n = 100
p = 512

FF = matrix(0, nrow=p, ncol=K)
f1 = 3
f2 = 1
FF[(p/8*1):(p/8*2),1] = f1
FF[(p/8*3):(p/8*4),2] = f2
FF[(p/8*5):(p/8*7),3] = f1

# l0 = 0.1
# l1 = 1
# L = cbind(c(rep(l1,n/3),rep(l0,n/3*2)),
#           c(rep(l0,n/3),rep(l1,n/3),rep(l0,n/3)),
#           c(rep(l0,n/3*2),rep(l1,n/3)))

L = matrix(rexp(n*K,1),nrow=n,ncol=K)

set.seed(12345)
y = matrix(rpois(n*p,tcrossprod(L,FF)),nrow=n)
```


```{r}
fit_init = init_poisson_nmf(y,F = FF,init.method = 'random')
out = fit_poisson_nmf(y,fit0=fit_init,update.factors = NULL)
```

Check if $F$ stays the same.

```{r}
plot(out$F[,1])
plot(out$F[,2])
plot(out$F[,3])
```


Check the fitted $L$

```{r}
plot(L[,1],out$L[,1])
plot(L[,2],out$L[,2])
plot(L[,3],out$L[,3])
```

Check normalized $L$:

```{r}
out2 = poisson2multinom(out)
L_normalized = t(apply(L,1,function(z){z/sum(z)}))
plot(L_normalized[,1],out2$L[,1])
abline(a=0,b=1)
plot(L_normalized[,2],out2$L[,2])
abline(a=0,b=1)
plot(L_normalized[,3],out2$L[,3])
abline(a=0,b=1)
```

## Summary

Suppose we have a known factor matrix $F\in R_+^{p\times K}$ where $K$ is the number of topics.

Then run the following to get the estiamted $L$.



```{r,eval=FALSE}
# FF: the known factor is FF, p by K
# y: y is the data matrix, n by p
fit_init = init_poisson_nmf(y,F = FF,init.method = 'random')
out = fit_poisson_nmf(y,fit0=fit_init,update.factors = NULL)
```

To get $L$, look at 

```{r,eval=FALSE}
out$L
```

To get normalized $L$ (Poisson to Multinomial vector, so that rows are sum to 1)

```{r,eval=FALSE}
out2 = poisson2multinom(out)
out2$L
```

