---
title: "normal-gamma autoregressive (NGAR) proces"
author: "Dongyue Xie"
date: "2021-06-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction


Simulate some examples from [NGAR](https://www.sciencedirect.com/science/article/pii/S0304407613002273).

\begin{equation}
    \beta_{t} = \sqrt{\frac{\psi_t}{\psi_{t-1}}}\phi\beta_{t-1}+\eta_{t},
\end{equation}

\begin{equation}
\begin{split}
      &\psi_t|\kappa_{t-1}\sim Gamma(\lambda+\kappa_{t-1},\frac{\lambda}{\mu(1-\rho)})
      \\
      &\kappa_{t-1}|\psi_{t-1} \sim Poisson(\frac{\rho\lambda\psi_{t-1}}{\mu(1-\rho)}).
\end{split}
\end{equation}
where $\beta_1|\psi_1\sim N(0,\psi_1)$ and $\psi_1\sim Gamma(\lambda,\lambda/\mu)$. The stationary distribution of $\psi_t$ is $Gamma(\lambda,\frac{\lambda}{\mu})$.


```{r}
GenNGAR = function(N = 128,lambda = 0.2,mu = 1,phi = 0.95,rho = 0.95){
  psi = c()
kappas = c()
beta = c()
psi[1] = rgamma(1,shape = lambda,rate = lambda/mu)
beta[1] = rnorm(1,0,sqrt(psi[1]))
for(t in 2:N){
  kappas[t] = rpois(1,rho*lambda*psi[t-1]/mu/(1-rho))
  psi[t] = rgamma(1,shape = lambda+kappas[t], rate = lambda/mu/(1-rho))
  beta[t] = sqrt(psi[t]/psi[t-1])*phi*beta[t-1]+rnorm(1,0,sqrt((1-phi^2)*psi[t]))
}
par(mfrow=c(3,1))
plot(kappas,type='l',main = paste('lambda=',lambda))
plot(psi,type='l')
plot(beta,type='l')
abline(h = 0,lty=2)
par(mfrow=c(1,1))
return(list(kappas=kappas,psi=psi,beta=beta))
}

temp = GenNGAR(lambda = 1)
temp = GenNGAR(lambda = 0.2,phi=0.95,rho=0.99)
```

The process can also be written as a product of normal AR(1) and Gamma AR(1)

$\beta_t = \sqrt{\psi_t}\delta_t$ where $\delta_t$ is generated from an AR(1) process with autocorrelation parameter $\phi$, i.e. $\delta_t = \phi\delta_{t-1}+e_t$ where $e_t\sim N(0,1-\phi^2)$. 

```{r}
GenNGAR2 = function(N = 128,lambda = 0.2,mu = 1,phi = 0.95,rho = 0.95){
  psi = c()
kappas = c()
delta = c()
psi[1] = rgamma(1,shape = lambda,rate = lambda/mu)
delta[1] = rnorm(1,0,1)
for(t in 2:N){
  kappas[t] = rpois(1,rho*lambda*psi[t-1]/mu/(1-rho))
  psi[t] = rgamma(1,shape = lambda+kappas[t], rate = lambda/mu/(1-rho))
  delta[t] = phi*delta[t-1]+rnorm(1,0,sqrt(1-phi^2))
}
beta = delta*sqrt(psi)
par(mfrow=c(3,1))
plot(delta,type='l',main = paste('lambda=',lambda))
plot(psi,type='l')
plot(beta,type='l')
abline(h = 0,lty=2)
par(mfrow=c(1,1))
return(list(kappas=kappas,psi=psi,beta=beta))
}

temp = GenNGAR2()
```






