---
title: "Outlier in Wavelet smoothing"
author: "Dongyue Xie"
date: "2020-02-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Investigate how the outlier will influecne the wavelet smoothing and how to automatically account for outliers.

Generate outliers from Huber's contamination model:

```{r}
set.seed(12345)
n = 256
x = seq(-2*pi,2*pi,length.out = n)
f = 1.5*sin(x) + sin(2*x)
e_z = rbinom(n,1,0.05)
e = e_z*rnorm(n,0,1) + (1-e_z)*rnorm(n,0,0.2)
y = f+e
library(smashr)
out = smash.gaus(y)
plot(y,col='grey80')
lines(out)
plot(y,col='grey80')
lines(ti.thresh(y))
```

Try robust scatter plot smoothing

```{r}
plot(y,col='grey80')
lines(runmed(y,11))
```


Let's see the outliers

```{r}
plot(x,y)
points(x[which(e_z==1)],y[which(e_z==1)],pch=16,col=4)
```

Running mad estimate of variance:

```{r}
library(wavethresh)
library(caTools)
J = log2(n)
x.w = wd(y, filter.number=1,family='DaubExPhase', type = "station")
win.size = 25
sigma.est = runmad(accessD(x.w, J - 1), win.size, endrule = "func")


plot(x,sigma.est,col='grey80')
points(x[which(e_z==1)],sigma.est[which(e_z==1)],pch=16,col=4)

plot(abs(y[which(e_z==1)] - f[which(e_z==1)]),sigma.est[which(e_z==1)])
abline(0,1)

# reduce window size

win.size = 10
sigma.est = runmad(accessD(x.w, J - 1), win.size, endrule = "func")


plot(x,sigma.est,col='grey80')
points(x[which(e_z==1)],sigma.est[which(e_z==1)],pch=16,col=4)

plot(abs(y[which(e_z==1)] - f[which(e_z==1)]),sigma.est[which(e_z==1)])
abline(0,1)

plot(y,col='grey80')
lines(smash.gaus(y,sigma.est))


# 

win.size = 3
sigma.est = runmad(accessD(x.w, J - 1), win.size, endrule = "func")


plot(x,sigma.est,col='grey80')
points(x[which(e_z==1)],sigma.est[which(e_z==1)],pch=16,col=4)

plot(abs(y[which(e_z==1)] - f[which(e_z==1)]),sigma.est[which(e_z==1)])
abline(0,1)

plot(x,y,col='grey80')
lines(x,smash.gaus(y,sigma.est))
lines(x,f,col=4)
```

## Summary 1

can not achieve robustness via change running MAD window size

Let's look at the finest level NDWT coefficients

```{r}
plot(x,accessD(x.w, J - 1))
points(x[which(e_z==1)],accessD(x.w, J - 1)[which(e_z==1)],pch=16,col=4)
```

Let's directly use the absolute deviation as the sigma.est. Much better....

```{r}
plot(y,col='grey80')
lines(smash.gaus(y,abs(accessD(x.w, J - 1)-median(accessD(x.w, J - 1)))))
```

```{r}
plot(y,col='grey80')
lines(ti.thresh(y,abs(accessD(x.w, J - 1)-median(accessD(x.w, J - 1)))))
```


Let's increase the number of outliers

```{r}
set.seed(12345)
e_z = rbinom(n,1,0.1)
e = e_z*rnorm(n,0,1) + (1-e_z)*rnorm(n,0,0.2)
y = f+e
x.w = wd(y, filter.number=1,family='DaubExPhase', type = "station")
```

```{r}
plot(y,col='grey80')
lines(smash.gaus(y,abs(accessD(x.w, J - 1)-median(accessD(x.w, J - 1)))))
```

```{r}
plot(y,col='grey80')
lines(ti.thresh(y,abs(accessD(x.w, J - 1)-median(accessD(x.w, J - 1)))))
```


Let's increase the variance of outliers

```{r}
set.seed(12345)
e_z = rbinom(n,1,0.1)
e = e_z*rnorm(n,0,3) + (1-e_z)*rnorm(n,0,0.2)
y = f+e
x.w = wd(y, filter.number=1,family='DaubExPhase', type = "station")
```

```{r}
plot(y,col='grey80')
lines(smash.gaus(y,abs(accessD(x.w, J - 1)-median(accessD(x.w, J - 1)))))
```

```{r}
plot(y,col='grey80')
lines(ti.thresh(y,abs(accessD(x.w, J - 1)-median(accessD(x.w, J - 1)))))
```

Let's increase the variance of outliers and variance of random errors

```{r}
set.seed(12345)
e_z = rbinom(n,1,0.1)
e = e_z*rnorm(n,0,4) + (1-e_z)*rnorm(n,0,1)
y = f+e
x.w = wd(y, filter.number=1,family='DaubExPhase', type = "station")
```

```{r}
plot(y,col='grey80')
lines(smash.gaus(y,abs(accessD(x.w, J - 1)-median(accessD(x.w, J - 1)))))
```

```{r}
plot(y,col='grey80')
lines(ti.thresh(y,abs(accessD(x.w, J - 1)-median(accessD(x.w, J - 1)))))
```


## Pseudo-data approach

http://stat.snu.ac.kr/heeseok/html/paper/robusttech.pdf

```{r}
smooth_outlier = function(y,maxiter=10,tol=1e-4){
  f_hat = ti.thresh(y)
  niter = 1
  while(niter<=maxiter){
    niter = niter + 1
    y_tilde = f_hat + tanh(y-f_hat)/2
    f_hat_new = ti.thresh(y_tilde)
    if(norm(f_hat-f_hat_new,'2')<=tol){
      break
    }else{
      f_hat = f_hat_new
    }
  }
  f_hat_new
}

plot(y,col='grey80')
lines(smooth_outlier(y))
```

Too slow. Does not work well.
