---
title: "Poisson mean scale invariant"
author: "Dongyue Xie"
date: "2019-11-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

See if scale-invariant version could handle large scale case.

```{r}
library(ebpm)
set.seed(12345)
pi0=0.8
lamda=c()
n=100
for(i in 1:100){
  idx = rbinom(1,1,pi0)
  if(idx){
    lamda[i] = rgamma(1,0.1,1)
  }else{
    lamda[i]=rgamma(1,50,1)
  }
}

lamda = sort(lamda)

hist(lamda,hist=20)

x = rpois(n,lamda)
s=1
fit = ebpm_gamma_mixture_single_scale(x,s)

plot(x/s,fit$posterior$mean,xlab = 'MLE',ylab = 'Posterior mean',pch=16)
abline(0,1)

plot(x/s,col='grey80',pch=16)
lines(fit$posterior$mean,type='p',pch=3,col=2)
legend('topleft',c('MLE','Posterior mean'),col=c(1,2),pch=c(16,3))
```

No shrinkage estimate when $x=0,1,2$.

How about $\exp(E\log\lambda)$?

```{r}
plot((x/s),col='grey80',pch=16)
lines(exp(fit$posterior$mean_log),type='p',pch=3,col=2)
legend('topleft',c('MLE','exp(Posterior mean log)'),col=c(1,2),pch=c(16,3))
```

It shrinks $E\log\lambda$ when $x$ is large.

Look at estimated values from 70 to 80

```{r}
(x/s)[70:80]
round(exp(fit$posterior$mean_log[70:80]),3)
```
Look at fitted prior

```{r}
fit$fitted_g
```


Let's increase scale to $s=100$.

```{r}
set.seed(12345)
s = 100
x = rpois(n,s*lamda)
fit = ebpm_gamma_mixture_single_scale(x,s)

plot(x/s,fit$posterior$mean,xlab = 'MLE',ylab = 'Posterior mean',pch=16)
abline(0,1)

plot(x/s,col='grey80',pch=16)
lines(fit$posterior$mean,type='p',pch=3,col=2)
legend('topleft',c('MLE','Posterior mean'),col=c(1,2),pch=c(16,3))
```

```{r}
(x/s)[60]
round(fit$posterior$mean[60],3)
```
No shrinkage estimate of $\lambda$.

How about $\exp(E\log\lambda)$?

```{r}
plot(x/s,exp(fit$posterior$mean_log),xlab = 'MLE',ylab = 'exp(Posterior mean log)',pch=16)
abline(0,1)


plot(x/s,col='grey80',pch=16)
lines(exp(fit$posterior$mean_log),type='p',pch=3,col=2)
legend('topleft',c('MLE','exp(Posterior mean log)'),col=c(1,2),pch=c(16,3))
```

```{r}
(x/s)[60]
round(exp(fit$posterior$mean_log)[60],3)
```

Some shrinkage, roughly $2\%-5\%$ for some values. Maybe this is the reason why 

Look at fitted prior

```{r}
fit$fitted_g
```
