---
title: "Poisson Mean"
author: "Dongyue Xie"
date: "2019-11-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Empirical Bayes Poisson Mean Problem

Empirical Bayes Poisson Mean problem dates back to mid-20th centrary. Robbins(1956) introduced empirical Bayes and took Poisson distribution as an example. 

Efron's paper Bayes, Oracle Bayes and Empirical Bayes is an interesting read. Corbet’s butterfly data mentioned in the paper is an very early example of Poisson mean and Poisson smoothing!

```{r}
library(ebmpmf)
set.seed(12345)
pi0=0.8
lamda=c()
n=100
for(i in 1:100){
  idx = rbinom(1,1,pi0)
  if(idx){
    lamda[i] = rgamma(1,0.1,1)
  }else{
    lamda[i]=rgamma(1,10,2)
  }
}

x = rpois(n,lamda)

# robbins method
marginal_x = function(xi,x){
  mean(x==xi)
}

robbins_est = function(x){
  est = c()
  for(i in 1:length(x)){
    est[i] = (x[i]+1)*marginal_x(x[i]+1,x)/marginal_x(x[i],x)
  }
  est
}

fit_robbins = robbins_est(x)

fit = ebpm_exp_mixture(x,scale='estimate',shape=0.1)

plot(x,fit$posterior$mean)
abline(0,1)

plot(x,col='grey80')
lines(fit$posterior$mean,type='p',pch=3,col=2)


x = x + runif(n,0,0.1)
fit = ebpm_exp_mixture(x,scale='estimate',shape=0.1)
plot(x,col='grey80')
lines(fit$posterior$mean,type='p',pch=3,col=2)

x = round(x)
fit = ebpm_exp_mixture(x,scale='estimate',shape=0.1)
plot(x,col='grey80')
lines(fit$posterior$mean,type='p',pch=3,col=2)

ss = 300
x = rpois(n,ss*lamda) + runif(n,0,0.1)

fit = ebpm_exp_mixture(x,ss,scale=c(1e-3,1e3),shape=0.1)
plot(x/ss,col='grey80')
lines(fit$posterior$mean,type='p',pch=3,col=2)

```




```{r}

library(gsl)
Gauss2F1b <- function(a,b,c,x){
    if(x>=0 & x<1){
        hyperg_2F1(a,b,c,x)
    }else{
        hyperg_2F1(a,c-b,c,1-1/(1-x))/(1-x)^a
    }
}


PoissonMean_DD = function(y,alpha,tau,gama){
  n=length(y)
  pmean_k = c()
  for(i in 1:n){
    pmean_k[i] = beta(1+alpha+1/2,y[i]+1/2)*Gauss2F1b(gama,1+alpha+1/2,y[i]+alpha+1+1,1-tau^2)/beta(alpha+1/2,y[i]+1/2)/Gauss2F1b(gama,alpha+1/2,y[i]+alpha+1,1-tau^2)
  }
  (1-pmean_k)*(y+alpha)
}

fit = PoissonMean_DD(x,1,0.5,0.5)

plot(x,col='grey80')
lines(fit,type='p',pch=3,col=2)
```
