---
title: "Redundant factors after NDWT and reflecting data"
author: "Dongyue Xie"
date: "2021-10-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Two possible reasons for causing redundant factors.

1. If using NDWT, the transformation introduces correlations.

2. To perform $YW = LFW+EW$, we need the number of columns of $Y$ to be a power of 2. If it's not, we usually extend the matrix $Y$ to the left and right, by mirroring the left and right data. After making $\#$col a power of 2, we can also reflect the whole data matrix by $cbind(Y,Y[,p:1])$ so that the starting and ending point are the same. This reflecting  is important for NDWT.

Now look at some simple examples.

## NDWT

```{r}
set.seed(12345)
n <- 50
p <- 256
K <- 3
snr <- 1

## Step 1: sample U, an orthogonal matrix
rand_semdef_sym_mat <- crossprod(matrix(runif(n * n), n, n))
rand_ortho_mat <- eigen(rand_semdef_sym_mat)$vector[, 1:K]
u_1 <- rand_ortho_mat[, 1]
u_2 <- rand_ortho_mat[, 2]
u_3 <- rand_ortho_mat[, 3]

f1 = c(rep(0,p/8), rep(1, p/4), rep(0, p/4), rep(0, p/4),rep(0,p/8))
f2 = c(rep(0,p/8), rep(0, p/4), rep(1, p/4), rep(0, p/4),rep(0,p/8))
f3 = c(rep(0,p/8), rep(0, p/4), rep(0, p/4), rep(1, p/4),rep(0,p/8))

L = cbind(u_1,u_2,u_3)
FF=cbind(f1,f2,f3)

plot(f1,type='l')
plot(f2,type='l')
plot(f3,type='l')

M = n / 3 * u_1 %*% t(f1) +
        n / 5 * u_2 %*% t(f2) +
        n / 6 * u_3 %*% t(f3)
v = var(c(M))/snr
X = M + matrix(rnorm(n*p,0,sqrt(v)),nrow=n,ncol=p)
```

First fit flash as baseline

```{r}
library(flashr)
fit.flash = flash(X,var_type='constant')
plot(fit.flash$ldf$f[,1],type='l')
plot(fit.flash$ldf$f[,2],type='l')
plot(fit.flash$ldf$f[,3],type='l')
```

Then fit smooth_flash. Haar wavelet, DWT.

```{r}
library(funflash)
fit.smooth_flash = smooth_flash(X,var_type = 'constant')
plot(fit.smooth_flash$ldf$f[,1],type='l')
plot(fit.smooth_flash$ldf$f[,2],type='l')
plot(fit.smooth_flash$ldf$f[,3],type='l')
```


Now DWT funflash.

```{r}
datax = funflash_set_data(X,reflect.data = FALSE,type='wavelet')
fit.dwt = funflash(datax,var.type = 'constant')
```

```{r}
fit.dwt$nfactors
plot(fit.dwt$ldf$f[,1],type='l')
plot(fit.dwt$ldf$f[,2],type='l')
plot(fit.dwt$ldf$f[,3],type='l')
```


Now NDWT funflash.

```{r}
datax = funflash_set_data(X,reflect.data = FALSE,type='station')
fit.ndwt = funflash(datax,var.type = 'constant')
```

```{r}
fit.ndwt$nfactors
plot(fit.ndwt$ldf$f[,1],type='l')
plot(fit.ndwt$ldf$f[,2],type='l')
plot(fit.ndwt$ldf$f[,3],type='l')
plot(fit.ndwt$ldf$f[,4],type='l')
plot(fit.ndwt$ldf$f[,5],type='l')
plot(fit.ndwt$ldf$f[,6],type='l')
```


## power of two


Now create an example that needs extending the data.

```{r}
set.seed(12345)
n <- 50
p <- 8*4*10
K <- 3
snr <- 1

## Step 1: sample U, an orthogonal matrix
rand_semdef_sym_mat <- crossprod(matrix(runif(n * n), n, n))
rand_ortho_mat <- eigen(rand_semdef_sym_mat)$vector[, 1:K]
u_1 <- rand_ortho_mat[, 1]
u_2 <- rand_ortho_mat[, 2]
u_3 <- rand_ortho_mat[, 3]

f1 = c(rep(0,p/8), rep(1, p/4), rep(0, p/4), rep(0, p/4),rep(0,p/8))
f2 = c(rep(0,p/8), rep(0, p/4), rep(1, p/4), rep(0, p/4),rep(0,p/8))
f3 = c(rep(0,p/8), rep(0, p/4), rep(0, p/4), rep(1, p/4),rep(0,p/8))

L = cbind(u_1,u_2,u_3)
FF=cbind(f1,f2,f3)

plot(f1,type='l')
plot(f2,type='l')
plot(f3,type='l')

M = n / 3 * u_1 %*% t(f1) +
        n / 5 * u_2 %*% t(f2) +
        n / 6 * u_3 %*% t(f3)
v = var(c(M))/snr
X = M + matrix(rnorm(n*p,0,sqrt(v)),nrow=n,ncol=p)
```

extend the data

```{r}
X.e = funflash:::reflect(X,F)
```


First fit flash as baseline

```{r}
idx = X.e$idx
fit.flash = flash(X.e$x,var_type='constant')
fit.flash$nfactors
plot(fit.flash$ldf$f[idx,1],type='l')
plot(fit.flash$ldf$f[idx,2],type='l')
plot(fit.flash$ldf$f[idx,3],type='l')
plot(fit.flash$ldf$f[idx,4],type='l')
plot(fit.flash$ldf$f[idx,5],type='l')
```

Then fit smooth_flash. Haar wavelet, DWT.

```{r}
fit.smooth_flash = smooth_flash(X.e$x,var_type = 'constant')
fit.smooth_flash$nfactors
plot(fit.smooth_flash$ldf$f[idx,1],type='l')
plot(fit.smooth_flash$ldf$f[idx,2],type='l')
plot(fit.smooth_flash$ldf$f[idx,3],type='l')
plot(fit.smooth_flash$ldf$f[idx,4],type='l')
plot(fit.smooth_flash$ldf$f[idx,5],type='l')
```

Now DWT funflash.

```{r}
datax = funflash_set_data(X.e$x,reflect.data = FALSE,type='wavelet')
fit.dwt = funflash(datax,var.type = 'constant')
```

```{r}
fit.dwt$nfactors
plot(fit.dwt$ldf$f[idx,1],type='l')
plot(fit.dwt$ldf$f[idx,2],type='l')
plot(fit.dwt$ldf$f[idx,3],type='l')
plot(fit.dwt$ldf$f[idx,4],type='l')
plot(fit.dwt$ldf$f[idx,5],type='l')
```





