---
title: "reflect data before wavelet transformation"
author: "Dongyue Xie"
date: "2021-10-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Do we need to reflect data to make the starting and end point the same when doing DWT and NDWT?

```{r}
n = 2^10
mu = c(rep(0,n/2),rep(10,n/2))
x = mu + rnorm(n)
```

First, we try DWT, no reflecting.

```{r}
library(funflash)
fit = smash_dwt(x,1,filter.number = 1,family = 'DaubExPhase',ebnm_fn = 'ebnm_ash')
plot(x,col='grey80')
lines(fit$mu.est)
```

Now with reflecting.

```{r}
library(smashr)
x.r = smashr::reflect(x)
fit = smash_dwt(x.r$x,1,filter.number = 1,family = 'DaubExPhase',ebnm_fn='ebnm_ash')
plot(x,col='grey80')
lines(fit$mu.est[x.r$idx])
```


How about other wavelet that (does not fit this datastet)?

DWT, no reflecting.

```{r}
fit = smash_dwt(x,1,filter.number = 10,family = 'DaubLeAsymm',ebnm_fn = 'ebnm_ash')
plot(x,col='grey80')
lines(fit$mu.est)
```

With reflecting.

```{r}
fit = smash_dwt(x.r$x,1,filter.number = 10,family = 'DaubLeAsymm',ebnm_fn='ebnm_ash')
plot(x,col='grey80')
lines(fit$mu.est[x.r$idx])
```


Then we try NDWT. No reflecting

```{r}
fit = ti.thresh(x,1)
plot(x,col='grey80')
lines(fit)
```

Now with reflecting.

```{r}
fit = ti.thresh(x.r$x,1)
plot(x,col='grey80')
lines(fit[x.r$idx])
```

## Conclusion

If using NDWT, we need to reflect the data. 

If using DWT and non-Haar basis, it seems better to reflect data.


