---
title: "Smoothed Matrix Factorization using wavelet and EBMF"
author: "Dongyue Xie"
output: workflowr::wflow_html
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## Introduction

Consider a matrix factorization problem \[Y = LF+E,\]

where $Y$ is a $N\times p$ matrix, $L$ is a $N\times K$ matrix, $F$ is a $K\times p$ matrix and $E$ is a $N\times p$ matrix of residuals. 

We assume each row of $F$ is smooth or spatially-structured. Then each row of $Y$ is from a smooth function/curve with added noises. Matrix $L$ is assumed to be sparse.

The question is how to estimate $L$ and $F$.

This is very similar to functional principal component analysis, which considers finding weights and principal components of a collection of curves. A common approach is adding a roughness penalty of the weights to obtain smooth estimates. 

Here, we consider using a specific basis to represent the smooth curves - wavelet. Let $W$ be the discrete wavelet transformation(DWT) matrix. We perform wavelet decomnposition on both sides then \[YW=LFW+EW,\]
or 
\[\tilde{Y} = L\tilde{F}+\tilde{E}.\]

Now each row of $\tilde{F}$ is sparse. We can then apply penalized matrix factorization algorithm(mainly universal thresholding methods) to $\tilde{Y}$ to obtain sparse estimates of $L$ and $\tilde{F}$. Applying inverse DWT gives $\hat{F}$.

Applying empirical Bayes wavelet shrinkage methods needs extra steps to deal with each level separately. 

## Simulation

In this simulation study, we choose EBMF framework and compare it with the wavelet approach.

```{r}
library(wavethresh)
source("code/wave_ebmf.R")
library(flashr)
# wavelet-based matrix factorization
#'@ y: observed matrix
#'@ k: number of factors
#'@ filter.number, family: wavelet type

WaveEBMF = function(y,k,filter.number = 1,family = 'DaubExPhase'){
  N=nrow(y)
  p=ncol(y)
  W = GenW(n=p,filter.number = filter.number,family = family)
  y_tilde = y%*%W
  f2 = flash(y_tilde,Kmax=k,var_type = 'constant',backfit = T,verbose=F)
  f2_fitted = flash_get_ldf(f2)
  f_hat = (W%*%f2_fitted$f)
  return(list(f=f_hat,l=f2_fitted$l))
}

```

### A single factor example

Simulate $N=200$ and $p=256$ under single-factor model 
\[l_i\sim \pi_0\delta_0+(1-\pi_0)\sum_{m=1}^5\frac{1}{5}N(0,\sigma^2_m)\]

**Step function factor**

$f$ is a step function. 

```{r}
rmse = function(x1,x2){sqrt(mean((x1-x2)^2))}
set.seed(12345)
N = 200
p = 256
pi0 = 0.3
f = c(rep(2,p/4), rep(5, p/4), rep(6, p/4), rep(2, p/4))
l = c(rep(0,N*pi0),rnorm(N*(1-pi0)/5,0,sqrt(0.25)),
      rnorm(N*(1-pi0)/5,0,sqrt(0.5)),
      rnorm(N*(1-pi0)/5,0,sqrt(1)),
      rnorm(N*(1-pi0)/5,0,sqrt(2)),
      rnorm(N*(1-pi0)/5,0,sqrt(4)))
plot(l)
plot(f,type='l')

y = l%*%t(f)+matrix(rnorm(N*p,0,1),ncol=p)

# apply flash directly
f1 = flash(y,var_type = 'by_row',verbose=T)

# apply wavelet transform

# use Haar wavelet

f2 = wave_ebmf(y)

paste('RMSE of flash estimate:',round(rmse(f1$fitted_values,l%*%t(f)),5))
paste('RMSE of wave_ebmf estimate:',round(rmse(f2$fitted_values,l%*%t(f)),5))

plot(f1$ldf$f,col = 2,type='l',xlab='',ylab='',main='flash')
plot(f2$ldf$f,col = 2,type='l',xlab='',ylab='',main='wave flash')

```

**HeavySine function factor**

```{r}
f=DJ.EX(p,signal = 2)$heavi
y = l%*%t(f)+matrix(rnorm(N*p,0,1),ncol=p)

# apply flash directly
f1 = flash(y,var_type = 'by_row')

# apply wavelet transform

# use symmlet10

f2 = wave_ebmf(y)

paste('RMSE of flash estimate:',round(rmse(f1$fitted_values,l%*%t(f)),5))
paste('RMSE of wave_ebmf estimate:',round(rmse(f2$fitted_values,l%*%t(f)),5))

plot(f1$ldf$f,col = 2,type='l',xlab='',ylab='',main='flash')
plot(f2$ldf$f,col = 2,type='l',xlab='',ylab='',main='wave flash')
```


**Spike function factor**

```{r}
spike.f = function(x) (0.75 * exp(-500 * (x - 0.23)^2) + 1.5 * exp(-2000 * (x - 0.33)^2) + 3 * exp(-8000 * (x - 0.47)^2) + 2.25 * exp(-16000 * 
    (x - 0.69)^2) + 0.5 * exp(-32000 * (x - 0.83)^2))

t = 1:p/p
f = 2*spike.f(t)

y = l%*%t(f)+matrix(rnorm(N*p,0,1),ncol=p)

f1 = flash(y,var_type = 'by_row')

# apply wavelet transform

# use symmlet10

f2 = wave_ebmf(y)

paste('RMSE of flash estimate:',round(rmse(f1$fitted_values,l%*%t(f)),5))
paste('RMSE of wave_ebmf estimate:',round(rmse(f2$fitted_values,l%*%t(f)),5))

plot(f1$ldf$f,col = 2,type='l',xlab='',ylab='',main='flash')
plot(f2$ldf$f,col = 2,type='l',xlab='',ylab='',main='wave flash')
```



### Three factors example

Simulate $N=200$ and $p=256$ under the factor model 
\[l_i\sim \pi_0\delta_0+(1-\pi_0)\sum_{m=1}^5\frac{1}{5}N(0,\sigma^2_m)\]

We set $K=3$ and three factors are step function, Heavysine and spike functions.

```{r}
K=3
set.seed(12345)
l1 = c(rep(0,N*pi0),rnorm(N*(1-pi0)/5,0,sqrt(0.25)),
      rnorm(N*(1-pi0)/5,0,sqrt(0.5)),
      rnorm(N*(1-pi0)/5,0,sqrt(1)),
      rnorm(N*(1-pi0)/5,0,sqrt(2)),
      rnorm(N*(1-pi0)/5,0,sqrt(4)))
l1 = l1[sample(1:N)]
l2 = c(rep(0,N*pi0),rnorm(N*(1-pi0)/5,0,sqrt(0.25)),
      rnorm(N*(1-pi0)/5,0,sqrt(0.5)),
      rnorm(N*(1-pi0)/5,0,sqrt(1)),
      rnorm(N*(1-pi0)/5,0,sqrt(2)),
      rnorm(N*(1-pi0)/5,0,sqrt(4)))
l2 = l2[sample(1:N)]
l3 = c(rep(0,N*pi0),rnorm(N*(1-pi0)/5,0,sqrt(0.25)),
      rnorm(N*(1-pi0)/5,0,sqrt(0.5)),
      rnorm(N*(1-pi0)/5,0,sqrt(1)),
      rnorm(N*(1-pi0)/5,0,sqrt(2)),
      rnorm(N*(1-pi0)/5,0,sqrt(4)))
l3 = l3[sample(1:N)]
L = cbind(l1,l2,l3)

f_1 = c(rep(2,p/4), rep(5, p/4), rep(6, p/4), rep(2, p/4))
f_2 = DJ.EX(p,signal = 2)$heavi
f_3 = 2*spike.f(t)
FF = rbind(f_1,f_2,f_3)
E = matrix(rnorm(N*p,0,1),ncol=p)
Y = L%*%FF + E

# apply flash directly
f1 = flash(Y,var_type = 'by_row')

# apply wavelet transform

# use symmlet10

f2 = wave_ebmf(Y)

plot(f1$ldf$f[,1],type='l')
plot(f1$ldf$f[,2],type='l')
plot(f1$ldf$f[,3],type='l')

plot(f2$ldf$f[,1],type='l')
plot(f2$ldf$f[,2],type='l')
plot(f2$ldf$f[,3],type='l')

```




```{r,eval=F,include=F}
# How about use Haar wavelet?
f2 = WaveEBMF(Y,k=3,filter.number = 1,family = 'DaubExPhase')


plot(-f1$f[,1],col = 2,type='l',xlab='',ylab='',main='Estimate of factor 1')
lines(f_1/norm(f_1,'2'),col='grey80',type='p')
lines(-f2$f[,1],col=4)
legend('topright',c('true f','flash','Waveflash'),col=c('grey80',2,4),lty=c(1,1,1))

plot(-f1$f[,2],col = 2,type='l',xlab='',ylab='',main='Estimate of factor 1')
lines(f_2/norm(f_2,'2'),col='grey80',type='p')
lines(-f2$f[,2],col=4)
legend('topright',c('true f','flash','Waveflash'),col=c('grey80',2,4),lty=c(1,1,1))

plot(f1$f[,3],col = 2,type='l',xlab='',ylab='',main='Estimate of factor 1')
lines(f_3/norm(f_3,'2'),col='grey80',type='p')
lines(f2$f[,3],col=4)
legend('topright',c('true f','flash','Waveflash'),col=c('grey80',2,4),lty=c(1,1,1))

### How about change the order of three factors?

FF = rbind(f_3,f_2,f_1)
Y = L%*%FF + E

# apply flash directly
f1 = flash(Y,Kmax=3,var_type = 'constant',backfit = F,verbose=F)
f1 = flash_get_ldf(f1)
f2 = WaveEBMF(Y,k=3,filter.number = 1,family = 'DaubExPhase')


plot(f1$f[,1],col = 2,type='l',xlab='',ylab='',main='Estimate of factor 1')
lines(f_1/norm(f_1,'2'),col='grey80',type='p')
lines(f2$f[,1],col=4)
legend('topright',c('true f','flash','Waveflash'),col=c('grey80',2,4),lty=c(1,1,1))

plot(-f1$f[,2],col = 2,type='l',xlab='',ylab='',main='Estimate of factor 1')
lines(f_2/norm(f_2,'2'),col='grey80',type='p')
lines(-f2$f[,2],col=4)
legend('topright',c('true f','flash','Waveflash'),col=c('grey80',2,4),lty=c(1,1,1))

plot(-f1$f[,3],col = 2,type='l',xlab='',ylab='',main='Estimate of factor 1')
lines(f_3/norm(f_3,'2'),col='grey80',type='p')
lines(-f2$f[,3],col=4)
legend('topright',c('true f','flash','Waveflash'),col=c('grey80',2,4),lty=c(1,1,1))
```

