---
title: "Apply stm to GTEx data"
author: "DongyueXie"
date: "2019-12-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Gtex Data from Yang Li. Data consist of per-base read counts from "GeuvadisLCL", "GTEXAdipose", "WholeBlood". 

Gene: RPS13. It has 6 exons and 10 splicing variants. 

```{r}
load('~/RPS13fit.RData')
library(stm)
```

## Full Data

Fit stm to the full dataset with $K=3$.

```{r}
lf = poisson2multinom(t(fit_stm_bmsm_RPS13$qf$Ef),fit_stm_bmsm_RPS13$ql$El)
for(i in 1:3){
  barplot(lf$L[,i],main=paste0('Estimate of loading ',i),xlab='',ylab = '',ylim = c(0,max(lf$L)),col=c(rep(2,88),rep(3,226),rep(4,927)))
}
for(i in 1:3){
  barplot(lf$FF[,i],main=paste0('Estimate of factor ',i),xlab='',ylab = '',ylim = c(0,max(lf$FF)))
}
```

## Remove WholeBlood

Fit stm to the dataset removing "WholeBlood" tissue with $K=3$.

```{r}
lf = poisson2multinom(t(fit_stm_bmsm_RPS13_2$qf$Ef),fit_stm_bmsm_RPS13_2$ql$El)
plot(lf$FF[,1],ylim=range(lf$FF),type='l',xlab = 'base',ylab='Intensity',main='Factors')
lines(lf$FF[,2],col=2)
lines(lf$FF[,3],col=4)


for(i in 1:3){
  barplot(lf$L[,i],main=paste0('Estimated loading ',i),ylab = '',ylim = c(0,max(lf$L)),col=c(rep(2,88),rep(3,226)))
}
```


Fit Xing's stm method to the dataset removing "WholeBlood" tissue with $K=3$.

```{r}
plot(xing_rps13$phi[1,],type='l',ylim = range(xing_rps13$phi),xlab = 'base',ylab='Intensity',main='Factors')
lines(xing_rps13$phi[2,],type='l',col=2)
lines(xing_rps13$phi[3,],type='l',col=4)
```


Fit stm to the dataset removing "WholeBlood" tissue with $K=10$.

```{r}
lf = poisson2multinom(t(fit_stm_bmsm_RPS13_3$qf$Ef),fit_stm_bmsm_RPS13_3$ql$El)


for(i in 1:10){
  barplot(lf$L[,i],main=paste0('Estimate of loading ',i),xlab='',ylab = '',ylim = c(0,max(lf$L)),col=c(rep(2,88),rep(3,226)))
}
for(i in 1:10){
  barplot(lf$FF[,i],main=paste0('Estimate of factor ',i),xlab='',ylab = '',ylim = c(0,max(lf$FF)))
}
```

## GTEx adipose tissue

Fit stm to the dataset using only adipose tissue with $K=5$.

```{r}
lf = poisson2multinom(t(fit_stm_bmsm_RPS13_4$qf$Ef),fit_stm_bmsm_RPS13_4$ql$El)
plot(lf$FF[,1],ylim=range(lf$FF),type='l',xlab = 'base',ylab='Intensity',main='Factors')
lines(lf$FF[,2],col=2)
lines(lf$FF[,3],col=4)
lines(lf$FF[,4],col=3)
lines(lf$FF[,5],col=6)
```

```{r}
for(i in 1:5){
  barplot(lf$L[,i],main=paste0('Estimate of loading ',i),xlab='',ylab = '',ylim = c(0,max(lf$L)))
}
for(i in 1:5){
  barplot(lf$FF[,i],main=paste0('Estimate of factor ',i),xlab='',ylab = '',ylim = c(0,max(lf$FF)))
}
```


Fit stm with smashgen using only adipose tissue with $K=3$.

```{r}
lf = poisson2multinom(t(fit_stm_sg_RPS13_4$qf$Ef),fit_stm_sg_RPS13_4$ql$El)
plot(lf$FF[,1],ylim=range(lf$FF),type='l',xlab = 'base',ylab='Intensity',main='Factors')
lines(lf$FF[,2],col=2)
lines(lf$FF[,3],col=4)
```


```{r}
for(i in 1:3){
  barplot(lf$L[,i],main=paste0('Estimate of loading ',i),xlab='',ylab = '',ylim = c(0,max(lf$L)))
}
for(i in 1:3){
  barplot(lf$FF[,i],main=paste0('Estimate of factor ',i),xlab='',ylab = '',ylim = c(0,max(lf$FF)))
}
```

Estimated Nugget effect:

```{r}
fit_stm_sg_RPS13_4$nugget$nugget_f
```
