---
title: "waveflash structure fix"
author: "Dongyue Xie"
date: "2022-05-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width=9,fig.height=3)
```

## Introduction

Previously we have found a problem with waveflash that it does not reveal structuresd very well. See an example [here](redundant_factors_reflect.html)

Later we found it's due to the missing energy term and the initialization(svd) without energy term in the wave coefficient space is different from what it should be. See an analysis [here](waveflash_structure2.html)

Now I re-implemented the model and considered the energy term. Let's see if it fixes the issue.

```{r,fig.width=9,fig.height=3}
set.seed(12345)
n <- 50
p <- 256
K <- 3
snr <- 1

## Step 1: sample U, an orthogonal matrix
rand_semdef_sym_mat <- crossprod(matrix(runif(n * n), n, n))
rand_ortho_mat <- eigen(rand_semdef_sym_mat)$vector[, 1:K]
u_1 <- rand_ortho_mat[, 1]
u_2 <- rand_ortho_mat[, 2]
u_3 <- rand_ortho_mat[, 3]

f1 = c(rep(0,p/8), rep(1, p/4), rep(0, p/4), rep(0, p/4),rep(0,p/8))
f2 = c(rep(0,p/8), rep(0, p/4), rep(1, p/4), rep(0, p/4),rep(0,p/8))
f3 = c(rep(0,p/8), rep(0, p/4), rep(0, p/4), rep(1, p/4),rep(0,p/8))

f1 = f1/sqrt(sum(f1^2))
f2 = f2/sqrt(sum(f2^2))
f3 = f3/sqrt(sum(f3^2))

FF=cbind(f1,f2,f3)

par(mfrow=c(1,3))

plot(f1,type='l')
plot(f2,type='l')
plot(f3,type='l')

M = n / 3 * u_1 %*% t(f1) +
        n / 5 * u_2 %*% t(f2) +
        n / 6 * u_3 %*% t(f3)
v = var(c(M))/snr
X = M + matrix(rnorm(n*p,0,sqrt(v)),nrow=n,ncol=p)
```


First fit flash as baseline

```{r,fig.width=9,fig.height=3}
library(flashr)
fit.flash = flash(X,var_type='by_row')
par(mfrow=c(1,3))
plot(fit.flash$ldf$f[,1],type='l')
plot(fit.flash$ldf$f[,2],type='l')
plot(fit.flash$ldf$f[,3],type='l')
```

Then fit smooth_flash. Haar wavelet, DWT.

```{r}
library(funflash)
fit.smooth_flash = smooth_flash(X,var_type = 'by_row')
par(mfrow=c(1,3))
plot(fit.smooth_flash$ldf$f[,1],type='l')
plot(fit.smooth_flash$ldf$f[,2],type='l')
plot(fit.smooth_flash$ldf$f[,3],type='l')
```


Now DWT funflash.

```{r}
datax = funflash_set_data(X,reflect.data = FALSE,type='wavelet')
fit.dwt = funflash(datax,var.type = 'by_row',nullcheck = F,Kmax=3)
```

```{r}
fit.dwt$nfactors
par(mfrow=c(1,3))
plot(fit.dwt$ldf$f[,1],type='l')
plot(fit.dwt$ldf$f[,2],type='l')
plot(fit.dwt$ldf$f[,3],type='l')
```

Compare the loadings.

```{r}
par(mfrow=c(1,3))
plot(u_1)
plot(u_2)
plot(u_3)
```

```{r}
par(mfrow=c(1,3))
plot(fit.flash$ldf$l[,1])
plot(fit.flash$ldf$l[,2])
plot(fit.flash$ldf$l[,3])
```

```{r}
par(mfrow=c(1,3))
plot(fit.smooth_flash$ldf$l[,1])
plot(fit.smooth_flash$ldf$l[,2])
plot(fit.smooth_flash$ldf$l[,3])
```

```{r}
par(mfrow=c(1,3))
plot(fit.dwt$ldf$l[,1])
plot(fit.dwt$ldf$l[,2])
plot(fit.dwt$ldf$l[,3])
```
