---
title: "waveflash structure a detailed look"
author: "Dongyue Xie"
date: "2022-05-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Solve the issue

We look into why running flash on wavelet coefficients yields different factorization from flash. See [examples](redundant_factors_reflect.html) and [More examples](waveflash_structure.html)

Here we generate orthogonal loadings as the eigen-vectors of a random matrix whose entries are drawn iid from a uniform[0,1] distribution.

Factors are orthogonal step functions.

```{r}
set.seed(12345)
n <- 50
p <- 256
K <- 3
snr <- 1

## Step 1: sample U, an orthogonal matrix
rand_semdef_sym_mat <- crossprod(matrix(runif(n * n), n, n))
rand_ortho_mat <- eigen(rand_semdef_sym_mat)$vector[, 1:K]
u_1 <- rand_ortho_mat[, 1]
u_2 <- rand_ortho_mat[, 2]
u_3 <- rand_ortho_mat[, 3]

f1 = c(rep(0,p/8), rep(1, p/4), rep(0, p/4), rep(0, p/4),rep(0,p/8))
f2 = c(rep(0,p/8), rep(0, p/4), rep(1, p/4), rep(0, p/4),rep(0,p/8))
f3 = c(rep(0,p/8), rep(0, p/4), rep(0, p/4), rep(1, p/4),rep(0,p/8))

f1 = f1/sqrt(sum(f1^2))
f2 = f2/sqrt(sum(f2^2))
f3 = f3/sqrt(sum(f3^2))

FF=cbind(f1,f2,f3)

plot(f1,type='l')
plot(f2,type='l')
plot(f3,type='l')

M = n / 3 * u_1 %*% t(f1) +
        n / 5 * u_2 %*% t(f2) +
        n / 6 * u_3 %*% t(f3)
v = var(c(M))/snr
X = M + matrix(rnorm(n*p,0,sqrt(v)),nrow=n,ncol=p)
```


```{r}
library(funflash)
datax = funflash_set_data(X,reflect.data = FALSE,type='wavelet')
plot(datax$Y[1,])
plot(datax$Y[2,])
plot(datax$Y[3,])
```

Look at the initialization.

```{r}
init_fn = 'udv_si_svd'
init1 = funflash:::init_fl(datax$Y,init_fn)
plot(init1$v)
init1$d
```

Now transform the initialization back to data space and see if it is the same as the initialization directly on $Y$.

```{r}
library(wavethresh)
wd.temp = wd(rep(0,p),filter.number = 1,family = 'DaubExPhase')
wd.temp$D = init1$v
#wd.temp$C[length(wd.temp$C)] = 1
v = wr(wd.temp)
v = v/sqrt(sum(v^2))
plot(v,main='v transformed from wavespace')

init_Y = funflash:::init_fl(X,init_fn)
init_Y$d
plot(init_Y$v,main='v from dataspace')
```

They look different!

Or transform the v from data space to wavespace.

```{r}
plot(wd(init_Y$v,filter.number = 1,family = 'DaubExPhase')$D,main='Wv, v from dataspace')
```

Ok, they are differenct confirmed.

One reason is that when performing rank1 svd (the initialization) in the wavelet space, I did not include the summation term. Now I include the summation term then do the initialization.


```{r}
init2 = funflash:::init_fl(cbind(datax$Y,datax$energy),init_fn)
plot(init2$v)
init2$d
```


```{r}
wd.temp = wd(rep(0,p),filter.number = 1,family = 'DaubExPhase')
wd.temp$D = init2$v[1:(p-1)]
wd.temp$C[length(wd.temp$C)] = init2$v[p]
v = wr(wd.temp)
v = v/sqrt(sum(v^2))
plot(v,main='v transformed from wavespace, with energy term when svd1')
```

Now they look the same!

```{r}
plot(v,init_Y$v)
```

They are the same, confirmed.

We need to consider the energy term when running the method and the results then should improve.


## SVD on Y,YW

If the SVD of $Y$ is $Y = UDV^T$ then $YW^T = UDV^TW^T=UD(WV)^T$ where $W$ is othornormal. The matrix $WV$ is also othornormal and $WV$ can be regarded as performing DWT on each $v$(column of V).

Is this true? Check it. 

```{r}
svd_Y = svd(X)
plot(svd_Y$v[,1],type='l')
plot(svd_Y$v[,2],type='l')
plot(svd_Y$v[,3],type='l')
```

```{r}
W = GenW(p,1,'DaubExPhase')
XW = X%*%W

svd_YW = svd(XW)

plot(svd_YW$d)
```


```{r}
plot(W%*%svd_YW$v[,1],type='l')
plot(W%*%svd_YW$v[,2],type='l')
plot(W%*%svd_YW$v[,3],type='l')
```

Also check the loadings(U).

```{r}
plot(svd_Y$u[,1],svd_YW$u[,1],col='grey80')
abline(a=0,b=1)
```

```{r}
plot(svd_Y$u[,2],svd_YW$u[,2],col='grey80')
abline(a=0,b=1)
```

```{r}
plot(svd_Y$u[,3],svd_YW$u[,3],col='grey80')
abline(a=0,b=1)
```




